\vspace{-15mm}\fontsize{24pt}{10pt}\selectfont\textbf{A New Radial Time Projection Chamber for CLAS at JLab}

### 2017-01-16 eg6-letters/NIM/EG6_RTPC_NIM.tex
update the NIM paper

\begin{abstract}
A new Radial Time Projection Chamber (RTPC) was developed at Jefferson 
laboratory (JLab) to track low-energy nuclear recoils for the purpose of 
measuring nuclear exclusive channels, such as, coherent Deeply Virtual Compton 
Scattering (DVCS) and coherent production of mesons on $^4$He. In such 
processes, $^4$He remains intact and recoils as a whole in the final state. The 
CEBAF Large Acceptance Spectrometer (CLAS) at JLab has been used for similar 
studies on proton targets, however it cannot track recoils with momenta less 
than 250 MeV/c. In 2009, we carried out measurements on $^4$He target using
CLAS spectrometer, which was upgraded with a RTPC. The detector was positioned 
directly around the gaseous target, allowing a detection threshold of $~$125  
MeV/c for $^4$He. This article discuses the design of our RTPC, the methods we 
developed to calibrate it and its performances.
\end{abstract}

@@

\section{\label{sec:level1} Introduction}

The CEBAF Large Acceptance Spectrometer (CLAS)~\cite{CLASref} is installed in 
the Hall-B of the Thomas Jefferson National Accelerator Facility (TJNAF) in 
Newport News, Virginia, USA. The spectrometer is measuring charged particles as 
well as photons. The accelerator of the TJNAF provides a high power electron 
beam of energy up to 6 GeV and 100$\%$ duty cycle to the hall. The RTPC 
mentioned in this article has been used during a three months experimental run 
in 2009 with a longitudinally polarized electron beam of energy 6.064 GeV sent 
on a gaseous helium-4 target to measure asymmetries in Deep Virtual Compton 
Scattering (DVCS)~\cite{proposal} and exclusive meson production (Meson Spec 
Prop). In these experiments, CLAS detects the high energy charged particles and 
the final real photon, for the latter the coverage has been extended using the 
optional inner calorimeter (ref).

\begin{figure}[h]
\centering 
\includegraphics[scale=0.3]{fig/test_clas.png}
\caption{\small\sf Three dimensional representation of the primary CLAS setup represents the geometry and the relative position of each subsystem. From outside, the electromagnetic calorimeter (green), time-of-flight system (red), Cherenkov counters (purple), three drift chamber regions (blue) and the superconducting torus coils (yellow).} 
\label{fig:CLAS}
\end{figure}

CLAS is composed of several subdetectors and two magnets. They are designed to track charged particles at momentum greater than 250 MeV/c. Figure~\ref{fig:CLAS} shows the basic setup of CLAS, ordered with respect to their radial distance from the center:

\begin{itemize}
\item Three regions of Drift Chambers (DC) for tracking of charged particles~\cite{DCref}.
\item Superconducting toroidal magnet which enables the tracking by bending the trajectories.
\item Cherenvkov Counters (CC) to separate electrons from negative pions~\cite{CCref}.
\item Scintillation Counters (SC) to identify hadrons via measuring their Time-Of-Flight (TOF)~\cite{TOFref}.
\item Forward Electromagnetic Calorimeter (EC) to detect showering particles (electrons and photons) and neutrons~\cite{ECref}.
\end{itemize}

In addition to the normal CLAS setup, a forward electromagnetic calorimeter (the inner calorimeter) was installed to improve the photon detection acceptance. Moreover, a 5~T solenoid was placed around the target to shield the detectors close to the beam line -- mostly the inner calorimeter and the RTPC -- from Mollers electrons. The solenoid field is also used to calculate the momentum of the alpha particles from their curvature radius measured by the RTPC.

The first implementation of a RTPC in CLAS have been achieved for the BoNuS experiment~\cite{BONUS} in 200x. The success of this detector for the detection of slow protons motivated the development of this new RTPC. For the channel of interests of our experiments, the recoiled $^{4}He$ nuclei have momentum around (150 or 300 ???) MeV/c, the detection of these particles was the primary motivation for using such design.

This paper is organized in this way: section \ref{sec_design} details the design and the internal structure of the RTPC, while section \ref{sec_readout} points the properties of the readout system. The calibration strategies for such detectors will be discussed in section \ref{sec_calib}. Finally, performances are discussed in section \ref{sec_perf}.

\section{Design} \label{sec_design}

The RTPC is a 200~mm long with a 150~mm diameter time projection chamber. The ionized electrons are drifting outward perpendicularly to the beam direction. These electrons are then amplified by three layers of GEMs and detected on 4.45 by 5~mm pads. As for BoNuS, the RTPC has two detection parts but unlike BoNuS they are part of a single chamber. Each part has independent GEM amplification systems that cover around 80\% of the azimuthal angle. Under the effect of the electromagnetic field, the ionization electrons drift to the readout system on the external shell with a curved trajectory as can be seen in the figure~\ref{fig:RTPC_1_4}.

\begin{figure}[tb]
\centering
\includegraphics[scale=0.19]{fig/RTPC_exp.png}
\caption[]{\small\sf View of the RTPC ready to be inserted in the solenoid. } 
\label{fig:RTPC2}
\end{figure}

\begin{figure}[tb]
\centering
\includegraphics[scale=0.28]{fig/RTPC_1_all.png}
\caption[]{\small\sf Schematic drawing of the RTPC taken on a plane perpendicular to the beam line, see text for description of the elements.} 
\label{fig:RTPC_1_4}
\end{figure}

As seen in figure~\ref{fig:RTPC2}, the RTPC is composed from the beam axis to the exterior by:\\
\begin{itemize}
\item The target, filled with 5~atm $^4$He, has a 3~mm radius with Kapton walls only 27$\mu$m thick.
\item The first gas gap extends from 3 mm to 20 mm radial distance. It is filled with $^{4}He$ gas at 1~atm. During run, this region is swarmed by Moller electrons induced by the beam, such that filling this region with a light gas like $^{4}He$ minimizes secondary interactions. This region is surrounded by a 4 $\mu$m thick window made of aluminized Mylar and connected to ground. 
\item The gap region extends between 20~mm and 30~mm and is filled with a gas mixture of 80$\%$ Neon and 20$\%$ Dimethyl Ether (DME). This region is surrounded by a 4~$\mu$m thick window made of aluminized Mylar, which serves as the cathode.
\item The drift region is filled with the Ne-DME gas mixture, it extends from the cathode to the first Gas Electron Multiplier(GEM), 60 mm away from the central axis. The electric field in this region is around 500~V/cm (???) and perpendicular to the beam. Field cages are installed at both ends of the detector in this region. Progressive voltage depending on radial position is applied on these to maintain a uniform electric field perpendicular to the beam line at the edges of the RTPC.
\item The GEM system is composed of three GEMs located at radii, 60 mm, 63 mm and 66 mm, they amplify the electrons to make a measurable signal.
\item The readout board has internal radius of 69~mm and collects the charges, preamplifiers are plugged directly on its outer side and are transmitting the signal to the data acquisition electronics.
\end{itemize}
\begin{figure}[tb]
\centering
\includegraphics[scale=0.70]{fig/GEM_photo.jpg}
\caption[]{\small\sf A schematic layout of the GEMs system in one module of the RTPC.} 
\label{fig:GEMs}
\end{figure}

@@

\begin{figure}[tb]
\centering
\includegraphics[scale=0.35]{fig/RTPC_2.png}
\caption[]{\small\sf A schematic drawing of a $^{4}He$ track (in green) traversing the drift region, with the drift paths followed by the electrons (in black). } 
\label{fig:RTPC_signals}
\end{figure}

@@

\begin{figure}[tb]
\centering
\includegraphics[scale=0.35]{fig/B_MAP.png}
\caption[]{\small\sf Schematic draw in of the solenoid magnetic vectors inside the RTPC. In this draw, r is the radial distance from the target's axis and z is the longitudinal position along the RTPC with its center at zero.} 
\label{fig:B_MAP}
\end{figure}

\begin{figure}[tb]
\centering
\includegraphics[scale=0.40]{fig/TDC_profile.png}
\caption[]{\small\sf Time profile of the collected hits in one experimental run. } 
\label{fig:TDC_profile}
\end{figure}

\begin{figure}[tb]
\centering
\includegraphics[scale=0.4]{fig/RunNumber_61551_TDCmax_Zslice.png}
\caption[]{\small\sf Time profile distribution for the collected hits in one experimental run. } 
\label{fig:RunNumber_61551_TDCmax_Zslice}
\end{figure}

@@

\begin{figure}[tb]
\hspace*{-1.8cm}
\includegraphics[scale=0.28]{fig/Drift_run_number_1.png}
\caption[]{\small\sf $TDC_{Nbmax/2}$ versus the experimental run numbers (time). } 
\label{fig:Drift_run_number_1}
\end{figure}

>>>>

\begin{abstract}
A new Radial Time Projection Chamber (RTPC) was developed at Jefferson 
laboratory (JLab) to track low-energy nuclear recoils for the purpose of 
measuring nuclear exclusive channels, such as, coherent Deeply Virtual Compton 
Scattering (DVCS) and coherent production of mesons on $^4$He. In such 
processes, $^4$He nucleus remains intact and recoils as a whole in the final 
state. The CEBAF Large Acceptance Spectrometer (CLAS) at JLab has been used for 
similar studies on proton targets, however it cannot track recoils with momenta 
less than 250 MeV/c. In 2009, we carried out measurements on $^4$He target 
using CLAS spectrometer, which was upgraded with the RTPC during the CLAS-EG6 
experimental run. The detector was positioned directly around a gaseous target, 
allowing a detection threshold of $~$100  MeV/c for $^4$He. This article 
discuses the RTPC design, work principle, developed calibration methods and 
detector performance.
\end{abstract}

@@

\section{Introduction} \label{sec:level1}

The CEBAF Large Acceptance Spectrometer (CLAS)~\cite{CLASref} is installed in 
the Hall-B of the Thomas Jefferson National Accelerator Facility (TJNAF), 
Newport News, Virginia, USA. The accelerator of TJNAF provides, simultaneously, 
a high power electron beam of energy up to 6 GeV and 100$\%$ duty cycle to 
three experimental Halls (A, B, C). The CLAS spectrometer, located in Hall-B, 
is composed of several sub-detectors and two magnets. The spectrometer was 
designed to track charged particles of momentum greater than 250 MeV/c as well 
as detecting real photons. Figure~\ref{fig:CLAS} shows a three dimensional 
representation of the CLAS detector, ordered with respect to their radial 
distance from the beam line as:

\begin{itemize}
 \item Three regions of Drift Chambers (DC) for tracking of charged 
    particles~\cite{DCref}.
 \item Superconducting toroidal magnet which enables the tracking in the DC by 
    bending the trajectories.
 \item Cherenvkov Counters (CC) to help in separating electrons from negative 
    particles, such as pions ~\cite{CCref}.
 \item Scintillation Counters (SC) to identify hadrons via measuring their 
    Time-Of-Flight (TOF)~\cite{TOFref}.
 \item Forward Electromagnetic Calorimeter (EC) which measures the energy of 
    the showering particles (electrons and photons) and neutrons~\cite{ECref}.
\end{itemize}

\begin{figure}[tbp]
\centering \includegraphics[scale=0.3]{fig/test_clas.png}
\caption{A three dimensional representation of the primary CLAS setup 
   represents the geometry and the relative position of each subsystem. From 
   outside, the electromagnetic calorimeter (green), time-of-flight system 
   (red), Cherenkov counters (purple), three drift chamber regions (blue) and 
the superconducting torus coils (yellow).} \label{fig:CLAS}
\end{figure}

The DVCS process is the exclusive electroproduction of a real photon from a 
quark of a hadronic target ($H$), $eH \rightarrow e' H' \gamma$. In the basic 
setup of CLAS, the real photons are detected by the forward EC, which covers 
polar angles from 8$^{\circ}$ to 45$^{\circ}$. With a 6 GeV electron beam, the 
majority of the DVCS photons are produced at polar angles below 15$^{\circ}$, 
where the acceptance of the EC is poor. In CLAS-E1DVCS experiment (2005), CLAS 
was upgraded with the addition of an Inner Calorimeter (IC). The IC is 
constructed from 424 lead-tungstate (PbWO$_{4}$) crystals, covering polar 
angles between 5$^{\circ}$ and 15$^{\circ}$. Each crystal is 16 cm long ,
corresponding to 17 radiation lengths, with a 1.33~$\times$~1.33 cm$^2$ front 
surface and a 1.6~$\times$~1.6 cm$^2$ back surface. The achieved energy 
resolution is around 3 to 4$\%$ for photon energies between 2 GeV and 5 GeV and 
the angular resolution is between 3 to 5 mrad for the same energy range 
\cite{Hyon-suk}.

The developed RTPC has been used during a three months experimental run in 2009 
with a longitudinally polarized electron beam of energy 6.064 GeV projected 
onto a gaseous $^{4}$He target to measure asymmetries in Deep Virtual Compton 
Scattering (DVCS)~\cite{proposal} and exclusive meson production. Moreover, a 
5~T solenoid was placed around the RTPC-target system to shield the detectors 
close to the beam line from Mollers electrons. The solenoid field plays also a 
major rule in finding the momentum of the charged particles in the RTPC by 
bending their tracks curvature.

In this paper we report the progress in designing and calibrating a radial TPC 
that had been used with JLab-CLAS spectrometer. This paper is organized as 
follows: section \ref{prev_tpc} presents an overview of the TPCs history and 
previous usage, section \ref{sec_design} details the physics considerations for 
the RTPC design and its internal structure, while section \ref{sec_readout} 
points the properties of the read-out system. The calibration strategies for 
such a detector are discussed in section \ref{sec_calib}. Finally, the 
performance of the RTPC is described in section \ref{sec_perf}.

\section{TPC technology} \label{prev_tpc}
In the late 1970s, David R. Nygren presented the concept of the Time projection 
Chambers (TPCs) as a new tool to track charged particles based on combining 
electric and magnetic fields within a sensitive gas volume. The ionized 
electrons are collected by a position-sensitive electron collection system.  
Later in 1997, F.~Sauli introduced the Gas Electron Multiplier (GEM) as an 
effective gas-gain element to improve the tracking efficiency of the TPCs 
\ref{gem_sauli}. Currently, the TPCs are widely used in physics experiments, 
such as high-energy physics (ALEPH, DELPHI), heavy-ion physics (STAR, ALICE) 
and rare-event searches (XENON, NEXT).

The first implementation of a radial TPC with JLab-CLAS spectrometer has been 
performed by the BoNuS experiment \cite{BONUS} in 2005. The success of the 
BoNuS-RTPC in detecting the slow protons motivated the development of this new 
RTPC. For the channel of interests of our experiments, the recoiled $^{4}He$ 
nuclei have momentum per charge around 150 MeV/c, the detection of these 
particles was the primary motivation for designing this detector.

\section{CLAS-EG6 RTPC design} \label{sec_design}
The design of CLAS-EG6 RTPC has been chosen according to physics and 
experimental considerations. At 6 GeV incident electron beam energy, the recoil 
$^{4}He$ nuclei, from the coherent DVCS channel, have an average momentum (per 
charge) around 125 MeV/c, while the "standard" CLAS detects charged particles 
with a threshold of 250 MeV/c. On the experimental site, The chosen design has 
to fit within the available space of a 5 Tesla Solenoid magnet that surrounds 
the target \cite{Hyon-suk}. Figure \ref{fig:RTPC2} shows an experimental view 
of the RTPC before inserting it into the solenoid.

The CLAS-EG6 RTPC is a 200~mm long with a 150~mm diameter time projection 
chamber. While a charged particle traverses a gas-sensitive volume, the ionized 
electrons are released and drift outward perpendicularly to the beam direction.  
These electrons are then amplified by three layers of GEMs and detected on 4.45 
by 5~mm pads. As for CLAS-BoNuS, the EG6-RTPC has two detection parts but 
unlike BoNuS they are part of a single chamber. Each part has independent GEM 
amplification systems that cover around 80\% of the azimuthal angle. Under the 
effect of the electromagnetic field, the ionization electrons drift to the 
readout system on the external shell with a curved trajectory as can be seen in 
the figure~\ref{fig:RTPC_1_4}.

\begin{figure}[tb]
\centering
\includegraphics[scale=0.19]{fig/RTPC_exp.png}
\caption{View of the RTPC ready to be inserted in the solenoid. } 
\label{fig:RTPC2}
\end{figure}

\begin{figure}[tb]
\centering
\includegraphics[scale=0.28]{fig/RTPC_1_all.png}
\caption{Schematic drawing of the RTPC taken on a plane perpendicular to the 
beam line, see text for description of the elements.} \label{fig:RTPC_1_4}
\end{figure}

As seen in figure~\ref{fig:RTPC_1_4}, the RTPC has the following substructure, 
from the beam axis to the exterior::\\
\begin{itemize}
   \item The target extends along the RTPC's central z-axis, with a diameter of 
      6 mm. It is enclosed in a 27-$\mu$m-thick Kapton wall.

   \item The first gas gap extends from 3 mm to 20 mm radial distance. It is 
   filled with $^{4}He$ gas at 1~atm. During run, this region is swarmed by 
   Moller electrons induced by the beam, such that filling this region with a 
   light gas like $^{4}He$ minimizes secondary interactions. This region is 
   surrounded by a 4 $\mu$m thick window made of aluminized Mylar and connected 
to ground.

   \item The gap region extends between 20~mm and 30~mm and is filled with a 
      gas mixture of 80$\%$ Neon and 20$\%$ Dimethyl Ether (DME).  This region 
      is surrounded by a 4~$\mu$m thick window made of aluminized Mylar, which 
      serves as the cathode.

   \item The drift region is filled with the Ne-DME gas mixture, it extends 
      from the cathode to the first Gas Electron Multiplier(GEM), 60 mm away 
      from the central axis. The electric field in this region is around 
      500~V/cm (???) and perpendicular to the beam. Field cages are installed 
      at both ends of the detector in this region. Progressive voltage 
      depending on radial position is applied on these to maintain a uniform 
      electric field perpendicular to the beam line at the edges of the RTPC.

   \item The GEM system is composed of three GEMs located at radii, 60 mm, 63 
      mm and 66 mm, they amplify the electrons to make a measurable signal. In 
      this configuration, the first GEM layer represents the anode.

   \item The readout board has internal radius of 69~mm and collects the 
      charges, preamplifiers are plugged directly on its outer side and are 
      transmitting the signal to the data acquisition electronics.
\end{itemize}

\begin{figure}[tbp]
\centering
\includegraphics[scale=0.70]{fig/GEM_photo.jpg}
\caption{A schematic layout of the GEMs system in one module of the RTPC.} 
\label{fig:GEMs}
\end{figure}

@@

\begin{figure}[tb]
\centering
\includegraphics[scale=0.35]{fig/RTPC_2.png}
\caption[]{A schematic drawing of a $^{4}He$ track (in green) traversing the 
drift region, with the drift paths followed by the electrons (in black). } 
\label{fig:RTPC_signals}
\end{figure}

@@

\begin{figure}[tb]
   \centering
   \includegraphics[scale=0.35]{fig/B_MAP.png}
   \caption[]{Schematic draw in of the solenoid magnetic vectors inside the 
   RTPC. In this draw, r is the radial distance from the target's axis and z is 
the longitudinal position along the RTPC with its center at zero.} 
\label{fig:B_MAP}
\end{figure}

\begin{figure}[tb]
   \centering
   \includegraphics[scale=0.3]{fig/TDC_profile.png}
   \caption[]{Time profile of the collected hits in one experimental run.  } 
   \label{fig:TDC_profile}
\end{figure}

\begin{figure}[tb]
\centering
\includegraphics[scale=0.36]{fig/RunNumber_61551_TDCmax_Zslice.png}
\caption{Time profile distribution for the collected hits in one experimental 
run. } \label{fig:RunNumber_61551_TDCmax_Zslice}
\end{figure}

@@

\begin{figure}[tb]
\hspace*{-1.8cm}
\includegraphics[scale=0.26]{fig/Drift_run_number_1.png}
\caption{$TDC_{Nbmax/2}$ versus the experimental run numbers (time).  } 
\label{fig:Drift_run_number_1}
\end{figure}
